On the treatment of discordant detrital zircon U--Pb data

### 2020-12-07 paper.tex
Updated following Micha Dietze's suggestions

%% Journal abbreviations (please use the same for preprints and final revised papers)

% Advances in Geosciences (adgeo)
% Advances in Radio Science (ars)
% Advances in Science and Research (asr)
% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
% Annales Geophysicae (angeo)
% Archives Animal Breeding (aab)
% ASTRA Proceedings (ap)
% Atmospheric Chemistry and Physics (acp)
% Atmospheric Measurement Techniques (amt)
% Biogeosciences (bg)
% Climate of the Past (cp)
% DEUQUA Special Publications (deuquasp)
% Drinking Water Engineering and Science (dwes)
% Earth Surface Dynamics (esurf)
% Earth System Dynamics (esd)
% Earth System Science Data (essd)
% E&G Quaternary Science Journal (egqsj)
% European Journal of Mineralogy (ejm)
% Fossil Record (fr)
% Geochronology (gchron)
% Geographica Helvetica (gh)
% Geoscience Communication (gc)
% Geoscientific Instrumentation, Methods and Data Systems (gi)
% Geoscientific Model Development (gmd)
% History of Geo- and Space Sciences (hgss)
% Hydrology and Earth System Sciences (hess)
% Journal of Bone and Joint Infection (jbji)
% Journal of Micropalaeontology (jm)
% Journal of Sensors and Sensor Systems (jsss)
% Magnetic Resonance (mr)
% Mechanical Sciences (ms)
% Natural Hazards and Earth System Sciences (nhess)
% Nonlinear Processes in Geophysics (npg)
% Ocean Science (os)
% Primate Biology (pb)
% Proceedings of the International Association of Hydrological Sciences (piahs)
% Scientific Drilling (sd)
% SOIL (soil)
% Solid Earth (se)
% The Cryosphere (tc)
% Weather and Climate Dynamics (wcd)
% Web Ecology (we)
% Wind Energy Science (wes)

@@

%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.

%% If an author is deceased, please mark the respective author name(s) with a dagger, e.g. "\Author[2,$\dag$]{Anton}{Aman}", and add a further "\affil[$\dag$]{deceased, 1 July 2019}".

%% If authors contributed equally, please mark the respective author names with an asterisk, e.g. "\Author[2,*]{Anton}{Aman}" and "\Author[3,*]{Bradley}{Bman}" and add a further affiliation: "\affil[*]{These authors contributed equally to this work.}".

@@

%% These dates will be inserted by Copernicus Publications during the typesetting process.

@@

\begin{abstract}
  Detrital zircon U--Pb geochronology is a staple of sedimentary
  provenance analysis and crustal evolution studies. Constructing
  detrital age spectra is straightforward for concordant
  \textsuperscript{206}Pb/\textsuperscript{238}U- and
  \textsuperscript{207}Pb/\textsuperscript{206}Pb-compositions.  But
  unfortunately, many detrital U--Pb datasets contain a significant
  proportion of discordant analyses. This situation is expected to
  worsen in the future as a result of increasing analytical precision,
  which reveals ever smaller deviations from isotopic concordance.
  The analysis of discordant U--Pb data involves two largely arbitrary
  decisions.

@@

  Applying these six discordance filters to a 10,000-grain dataset of
  detrital zircon U--Pb compositions reveals that: (i) the relative
  age discordance filter tends to suppress the young ($<1.5$~Ga) age
  components in U--Pb age spectra, whilst inflating the older
  ($>1.5$~Ga) age components; (ii) the absolute age difference and
  Stacey-Kramers discordance filters are more likely to reject old
  grains and less likely to reject young ones; (iii) the p-value based
  discordance filter has the undesirable effect of biasing the results
  towards the least precise measurements; (iv) the logratio-based
  discordance filters are most strict for Proterozoic grains, and more
  lenient for Phanerozoic and Archaean age components; (v) of all the
  methods, the logratio distance to the concordia composition produces
  the best results, in the sense that it yields nearly identical
  \textsuperscript{206}Pb/\textsuperscript{238}U and
  \textsuperscript{207}Pb/\textsuperscript{206}Pb age spectra.  All
  the methods presented in this paper have been implemented in the
  \texttt{IsoplotR} toolbox for geochronology.
\end{abstract}

@@

Detrital zircon U--Pb geochronology is a widely used tool for
sedimentary provenance analysis, crustal evolution studies and maximum
depositional age estimation.  In all these applications, inferences
are based on the frequency distribution of U--Pb age \emph{estimates},
which are derived from isotopic ratio \emph{measurements}. Several
mathematical and statistical operations are required to extract an age
spectrum from a table of U--Pb isotope ratio measurements. There
exists a lack of consensus among the detrital zircon geochronology
community about some of these steps.  Two outstanding questions are:

\begin{enumerate}
\item Which age estimate to use? It is widely recognised that
  \textsuperscript{206}Pb/\textsuperscript{238}U age estimates offer
  the optimal accuracy and precision at the young end of the age
  spectrum, whereas the
  \textsuperscript{207}Pb/\textsuperscript{206}Pb method is better
  suited for older samples. However the cutoff between the two clocks
  varies between studies, with values ranging from 800~Ma to 1.5~Ga
  \citep{gehrels2011,spencer2016}.
\item How to treat discordant data? The most reliable age constraints
  are obtained from zircons whose
  \textsuperscript{206}Pb/\textsuperscript{238}U and
  \textsuperscript{207}Pb/\textsuperscript{206}Pb ages are
  statistically indistinguishable from each other. U--Pb compositions
  that fulfil this requirements are `concordant'. Those that violate
  it are `discordant'. It is not obvious how to quantify the
  discordance of the \textsuperscript{206}Pb/\textsuperscript{238}U
  and \textsuperscript{207}Pb/\textsuperscript{206}Pb systems. Most
  studies define discordance as the relative age difference between
  the \textsuperscript{206}Pb/\textsuperscript{238}U and
  \textsuperscript{207}Pb/\textsuperscript{206}Pb ages, but some
  advocate the use of statistical hypothesis tests and p-values to
  quantify discordance \citep{spencer2016}. And even when a
  discordance definition has been agreed upon, there are many ways to
  choose the discordance cutoff. For example, the relative age
  discordance threshold may vary between 10\% and 30\%
  \citep{gehrels2011}.
\end{enumerate}

This paper addresses both of these issues. Section~\ref{sec:whichage}
shows that single-grain concordia ages \citep{ludwig1998} can be used
to avoid the arbitrary cutoff between the
\textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb
methods. Section~\ref{sec:discordance1} compares and contrasts
existing discordance filters based on age disparity and p-values. It
shows that the p-value definition hurts both the accuracy and
precision of detrital geochronology. Section~\ref{sec:discordance2}
proposes three new ways to quantify discordance, based directly on
U--Pb compositions rather than the ages calculated therefrom. Two of
these new definitions are based on a logratio distance that is widely
used in compositional data analysis \citep{aitchison1986}. Finally,
Section~\ref{sec:application} applies the six discordance filters to a
compilation of detrital zircon U--Pb data. Although the true age
distribution of this dataset is unknowable, the results suggest that
the logratio based discordance filters produce the most accurate
results.

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{TW1500.pdf}
  \caption{Illustrative Tera-Wasserburg concordia diagram with a
      concordant and discordant measurement. $t_{68}$ marks the
      \textsuperscript{206}Pb/\textsuperscript{238}U age, $t_{76}$ the
      \textsuperscript{207}Pb/\textsuperscript{206}Pb age, and $t_{c}$
      the concordia age. Measurement~1 is concordant because its
      estimates for $t_{68}$, $t_{76}$ and $t_{c}$ are
      identical. Measurement~2 is discordant because the three
      estimates disagree. The concordia age is the most likely age
      given the analytical uncertainties. It falls between the other
      two age estimates, and offers the best analytical precision of
      the three.
  }
  \label{fig:concordia}
\end{figure}

@@

The single grain concordia age combines the chronometric power of the
\textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb systems. For young
($<$1~Ga) samples, the concordia age is nearly identical to the
\textsuperscript{206}Pb/\textsuperscript{238}U age. For old samples
($>$2~Ga) it approaches the
\textsuperscript{207}Pb/\textsuperscript{206}Pb age. For Proterozoic
samples, the concordia age gradually shifts form the
\textsuperscript{206}Pb/\textsuperscript{238}U to the
\textsuperscript{207}Pb/\textsuperscript{206}Pb age. Using concordia
ages removes the need for an arbitrary cutoff between the two
chronometers. An additional advantage is that the concordia age offers
better precision than the
\textsuperscript{206}Pb/\textsuperscript{238}U and the
\textsuperscript{207}Pb/\textsuperscript{206}Pb chronometer (or the
\textsuperscript{207}Pb/\textsuperscript{235}U for that matter)
(Figure~\ref{fig:precision}).

\begin{figure}[t]
  \includegraphics[width=8.3cm]{agerr.pdf}
  \caption{Predicted uncertainties of the
    \textsuperscript{207}Pb/\textsuperscript{235}U,
    \textsuperscript{206}Pb/\textsuperscript{238}U,
    \textsuperscript{207}Pb/\textsuperscript{206}Pb and concordia
    ages for a synthetic dataset with a constant uranium
    concentration. Dwell times and detector sensitivities were
    chosen so as to yield results that are similar to those obtained
    from real data. The concordia age (solid line) always offers the
    best precision. See the Appendix for further details about the
    calculations behind this figure.}
  \label{fig:precision}
\end{figure}

@@

Discordance may be caused by the addition of common Pb, the partial
removal of Pb during high grade metamorphism, or the mixing of
diachronous growth zones during micro-analysis. Most commonly,
discordance is defined in terms of the relative difference between the
\textsuperscript{206}Pb/\textsuperscript{238}U age and the
\textsuperscript{207}Pb/\textsuperscript{206}Pb age
\citep{gehrels2011}:
\begin{equation}
  d_r = 1 - t_{68}/t_{76}
  \label{eq:dr}
\end{equation}

@@

Detrital zircon U--Pb data can be filtered by removing all
measurements whose discordance values exceed a certain threshold
value. Typical cutoff values for $d_r$ are 10--30\%
\citep{gehrels2011}, whereas $d_p$ is generally set to 5\%
\citep{spencer2016}. Different discordance criteria produce different
U--Pb age spectra. For example, a relative age cutoff will
preferentially remove young grains whereas an absolute age cutoff is
more likely to remove old grains (Figure~\ref{fig:agediscordance}).

\begin{figure}[t]
  \includegraphics[width=8.3cm]{TW-option-12.pdf}
  \caption{Discordance filters using a 200~Ma absolute age cutoff
    (dashed lines) and a 20\% relative age cutoff (dotted
    lines). The Tera-Wasserburg concordia diagram is plotted in
    logarithmic space to provide a more balanced view of the old and
    young ends of the time scale.
  }
  \label{fig:agediscordance}
\end{figure}

The p-value definition affects grains differently depending on their
analytical precision \citep{nemchin2005}. For example, consider a
1.5~Ga zircon that is 1\% discordant. If this grain were analysed by
LA-ICP-MS with an analytical precision of 2\%, say, then it would pass
the chi-square test and be accepted as being concordant. However, if
that same grain were analysed by TIMS with a precision of 0.2\%, then
the p-value criterion would reject it as being discordant. It seems
fundamentally unfair that an imprecise analytical method would be
favoured over a precise one (Figure~\ref{fig:TIMSvsLAICPMS}). This is
a pertinent problem because technical innovations are increasing the
precision of all analytical approaches to U--Pb geochronology.  As
precision improves, so does the ability to detect even small degrees
of discordance. Using the p-value criterion, there may come a time
when no detrital zircon passes this filter.

@@

\begin{figure}
  \includegraphics[width=8.3cm]{TIMSvsICPMS.pdf}
  \caption{Application of the flawed p-value discordance criterion
    to two synthetic measurements by TIMS (left) and LA-ICP-MS
    (right).  The precise TIMS measurement is labelled as discordant
    even though it plots closer to the concordia line than the
    imprecise LA-ICP-MS measurement, which is labelled as
    concordant.
  }
  \label{fig:TIMSvsLAICPMS}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=8.3cm]{TW-option-linear-3.pdf}
    \caption{Using the \citet{stacey1975} common Pb model as a
      discordance criterion. This criterion assumes that the
      discordance is caused by linear mixing (hence, the linear scale
      of this Tera-Wasserburg plot) between radiogenic Pb
      (intersections of the mixing lines with concordia) and common Pb
      (intersection of the mixing lines with the vertical axis). The
      dashed line marks the 20\% ($= d_1/[d_1+d_2]$) discordance
      cutoff. This discordance filter, which must be applied
      \emph{before} making any actual common Pb correction, is more
      forgiving for young grains than it is for old grains. In this
      respect, it has the opposite effect of the relative age filter
      shown in Figure~\ref{fig:agediscordance}.
    }
    \label{fig:SK}
\end{figure}

@@

There exists a vast body of statistical literature detailing the
theoretical and practical advantages of logratio analysis. A deeper
discussion of this topic falls outside the scope of this paper, but
the interested reader is referred to \citet{aitchison1986} and
\citet{pawlowsky2015} for further information. The Euclidean distance
between logratios is also known as the `Aitchison distance'.  We can
redefine discordance as the Aitchison distance from the measured
logratios to the concordia line. We here introduce two ways to do so.
A first option is to simply measure the distance along a perpendicular
line to the concordia curve (Figure~\ref{fig:aitchison}):
\begin{equation}
  d_{a} = dx(t_{68})
  \sin\!\left(\arctan\!\left[ \frac{dy(t_{76})}{dx(t_{68})} \right]
  \right)
  \label{eq:dperp}
\end{equation}

@@

This definition produces a parallel band around the concordia line in
logarithmic Tera-Wasserburg space. In contrast with $d_r$, $d_t$,
$d_{sk}$, the $d_a$ criterion is less strict at both the young and old
extremes of the geological timescale, and more strict during the
Proterozoic Era, when the U--Pb method is most reliable
(Figure~\ref{fig:dperp}).

@@

\begin{figure}
  \includegraphics[width=8.3cm]{TW-option-4.pdf}
  \caption{The $d_a$ criterion produces a parallel band to the
    (logarithmic) Tera-Wasserburg concordia line. The dashed line is
    plotted at an Aitchison distance of 0.1 from concordia, marking
    a 10\% discordance cutoff. The bottom panel presents the age
    discordance along this line. This shows that the $d_a$ criterion
    is lenient for Phanerozoic and Archaean age estimates, and more
    strict for Proterozoic grains.
  }
  \label{fig:dperp}
\end{figure}

@@

The data come from field areas on all seven continents. They were
acquired by LA-ICP-MS, on either an Agilent 7700x (old samples) or an
Agilent 7900 (recent samples) instrument, which was coupled to a New
Wave NWR193 excimer laser. Analytical conditions varied slightly
between different samples but generally used a 25 or 35 micron laser
spot, Ple\v{s}ovice zircon as a primary standard \citep{slama2008}
and, in most cases, GJ-1 zircon as a secondary standard
\citep{jackson2004}. Data reduction was done with GLITTER
\citep{griffin2008}. The data were not subjected to any common Pb
correction or other filters, apart from a visual inspection to remove
the most extreme outliers ($<1\%$ of the data).

@@

\begin{figure}
  \includegraphics[width=8.3cm]{KDE.pdf}
  \caption{Four superimposed kernel density estimates (with 60~Ma
    bandwidth) for 10,000 unfiltered detrital zircon U--Pb
    dates. The \textsuperscript{207}Pb/\textsuperscript{235}U,
    \textsuperscript{206}Pb/\textsuperscript{238}U and concordia age
    spectra are similar.  However the KDE of the
    \textsuperscript{207}Pb/\textsuperscript{206}Pb data stands
    apart from the other three curves. It deviates both at the young
    end of the age spectrum (which it suppresses), and at the old
    end (which it inflates).
  }
  \label{fig:KDE}
\end{figure}

@@

\begin{figure*}[t]
  \includegraphics[width=12cm]{KDEs.pdf}
  \caption{Filtered U--Pb age spectra for the test data, removing the
    50\% most discordant grains according to the six discordance
    filters reviewed in this paper. Notice how the concordia distance
    filter ($d_c$, fifth panel) exhibits the smallest difference
    between the four dating methods (including
    \textsuperscript{207}Pb/\textsuperscript{206}Pb), wherease the
    p-value filter ($d_p$, sixth panel) generates the greatest
    differences between them.  This observation suggests that the
    $d_c$ criterion is the best and the $d_p$ criterion the worst of
    the six methods.
  }
  \label{fig:KDEs}
\end{figure*}

Figure~\ref{fig:KDEs} removed 50\% of the data, in order to emphasise
the differences between the six filters. In real applications, less
stringents discordance filters are usually applied. As mentioned in
the introduction, most current detrital zircon studies apply a
10\%--30\% relative age cutoff.  Using the test data, we can evaluate
the equivalent values for the $d_t$, $d_t$, $d_{sk}$, $d_a$ and $d_c$
criteria (Table~\ref{tab:conversiontable}). For example, a relative
age filter of 10\% removes the same fraction of the test data as an
absolute age filter with $d_t=83.5$~Ma, a Stacey-Kramers filter with
$d_{sk}=0.381$\%, a perpendicular Aitchison filter with $d_a=3.61$\%,
and a concordia distance filter with $d_c=3.67$\%.

The p-value discordance filter has been omitted from this comparison
for two reasons. First, the use of this filter is discouraged for
reasons that have been discussed before. Second, the p-value cutoffs
that are equivalent to any given relative age difference are highly
lab-dependent, with precise equipment requiring different
$d_p$-cutoffs than imprecise instruments. The other five discordance
filters are more universally applicable. So using a different set of
test data should only make a relatively minor difference to the values
in Table~\ref{tab:conversiontable}.

>>>>

@@

@@

@@

\begin{abstract}
  Detrital zircon U--Pb geochronology is a staple of sedimentary
  provenance analysis and crustal evolution studies. Constructing
  detrital age spectra is straightforward for concordant
  \textsuperscript{206}Pb/\textsuperscript{238}U- and
  \textsuperscript{207}Pb/\textsuperscript{206}Pb-compositions.  But
  unfortunately, many detrital U--Pb datasets contain a significant
  proportion of discordant analyses. This paper investigates two
  decisions that must be made when analysing such discordant U--Pb
  data.

@@

  Applying these six discordance filters to a 10,000-grain dataset of
  detrital zircon U--Pb compositions reveals that: (i) the relative
  age discordance filter tends to suppress the young age components in
  U--Pb age spectra, whilst inflating the older age components; (ii)
  the Stacey-Kramers discordance filter is more likely to reject old
  grains and less likely to reject young ones; (iii) the p-value based
  discordance filter has the undesirable effect of biasing the results
  towards the least precise measurements; (iv) the logratio-based
  discordance filters are most strict for Proterozoic grains, and more
  lenient for Phanerozoic and Archaean age components; (v) of all the
  methods, the logratio distance to the concordia composition produces
  the best results, in the sense that it yields nearly identical
  \textsuperscript{206}Pb/\textsuperscript{238}U and
  \textsuperscript{207}Pb/\textsuperscript{206}Pb age spectra without
  introducing an age bias.  All the methods presented in this paper
  have been implemented in the \texttt{IsoplotR} toolbox for
  geochronology.
\end{abstract}

@@

The U--Pb method consists of two paired decay systems, in which two
isotopes of the same radioactive parent (\textsuperscript{238}U and
\textsuperscript{235}U) decay to two isotopes of the same radiogenic
daughter (\textsuperscript{206}Pb and \textsuperscript{207}Pb,
respectively). This paired decay system provides a powerful internal
consistency check for the method, which is absent from other
chronometers. By `double dating' samples with the
\textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{235}U methods (or,
equivalently, the \textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb-methods) it is
possible to verify whether the isotopic system is free of primary or
secondary disturbances. The most reliable age constraints are obtained
from samples whose \textsuperscript{206}Pb/\textsuperscript{238}U,
\textsuperscript{207}Pb/\textsuperscript{235}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb ages are statistically
indistinguishable from each other. U--Pb compositions that fulfil this
requirements are `concordant'. Those that violate it are `discordant'.

Discordance can be caused by a number of mechanisms, including: (a)
the presence of non-radiogenic (`common') lead; (b) initial
disequilibrium between the short-lived nuclides of the
\textsuperscript{238}U--\textsuperscript{206}Pb and
\textsuperscript{235}U--\textsuperscript{207}Pb decay chains; (c)
partial loss of radiogenic lead during high grade metamorphism; and
(d) mixing of different age domains during micro-analysis
\citep{schoene2014}. These complicating effects can often be diagnosed
and remediated when multiple cogenetic crystals are available from the
same sample. If the aliquots form an isochron (or `discordia') line in
\textsuperscript{238}U-\textsuperscript{204,6,7}Pb space, then this
line can be used to recover robust chronologies from discordant data
\citep{ludwig1998}.

Unfortunately, this procedure is rarely or never possible for detrital
samples, in which crystals of datable minerals are not guaranteed to
be cogenetic.  Without a universal mechanism to identify the cause of
U--Pb discordance and remove its effects, detrital geochronologists
have no choice but to accept some discordant analyses and somehow
incorporate them into their age spectra.  There exists a lack of
consensus among the detrital zircon geochronology community on how to
do this.  Two outstanding questions are:

\begin{enumerate}
\item Which age estimate to use? It is widely recognised that
  \textsuperscript{206}Pb/\textsuperscript{238}U age estimates offer
  the optimal accuracy and precision at the young end of the age
  spectrum, whereas the
  \textsuperscript{207}Pb/\textsuperscript{206}Pb method is better
  suited for older samples. However the cutoff between the two clocks
  varies between studies, with values ranging from 800~Ma to 1.5~Ga
  \citep{gehrels2011,spencer2016}.
\item How to quantify discordance?  Most studies define discordance as
  the relative age difference between the
  \textsuperscript{206}Pb/\textsuperscript{238}U and
  \textsuperscript{207}Pb/\textsuperscript{206}Pb ages, but some
  advocate the use of statistical hypothesis tests and p-values to
  quantify discordance \citep{spencer2016}. And even when a
  discordance definition has been agreed upon, there are many ways to
  choose the discordance cutoff. For example, the relative age
  discordance threshold may vary between 10\% and 30\%
  \citep{gehrels2011}.
\end{enumerate}

This paper addresses both of these issues. Section~\ref{sec:whichage}
advocates the use of single-grain concordia ages \citep{ludwig1998} as
a way to avoid the arbitrary cutoff between the
\textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb methods. Although
previous workers have argued for the use of single-grain concordia
ages before \citep[see][for a recent example]{zimmermann2018}, this
study uses a semi-analytical model, rather than purely empirical
arguments, to demonstrate the superior precision of this hybrid
chronometer.

Section~\ref{sec:discordance1} compares and contrasts existing
discordance filters based on age disparity and p-values. It shows that
the relative age definition strongly favours older samples over young
ones, and that the p-value definition, which has gained popularity in
recent years, hurts both the accuracy and precision of detrital
geochronology. The age disparity and p-value definitions are heuristic
by nature and are not based on firm statistical or geological
arguments.  Although they are essentially the only two definitions of
discordance that are used in detrital geochronology today, they are by
no means the only two possible options.

Section~\ref{sec:discordance2} addresses the inherent biases of the
existing discordance definitions by proposing three new definitions,
which are based directly on U--Pb compositions rather than on the ages
calculated therefrom. The first new definition assumes that the
discordance is caused by the presence of common lead. The other two
new definitions treat U--Pb discordance as a compositional data
problem \citep[\textit{sensu}][]{aitchison1986}.  Isotope ratios are
strictly positive quantities and log-contrasts are the `natural' way
to quantify `distances' between them.  Section~\ref{sec:discordance2}
introduces two logratio definitions of discordance, ignoring and
accounting for analytical uncertainty, respectively.

Although the new definitions are arguably more attractive than the old
ones from a theoretical point of view, this does not guarantee that
they produce more sensible results. To test their performance on real
data, Section~\ref{sec:application} applies all six discordance
filters to a compilation of detrital zircon U--Pb data. Although the
true age distribution of this dataset is unknowable, the results
suggest that the logratio based discordance filters produce the most
accurate, and most easily interpretable results. The p-value
definition fares the worst and is not recommended.

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{TW1500.pdf}
  \caption{Illustrative Tera-Wasserburg concordia diagram with a
    concordant and discordant measurement. $t_{68}$ marks the
    \textsuperscript{206}Pb/\textsuperscript{238}U age, $t_{76}$ the
    \textsuperscript{207}Pb/\textsuperscript{206}Pb age, and $t_{c}$
    the concordia age. Measurement~1 is concordant because its
    estimates for $t_{68}$, $t_{76}$ and $t_{c}$ are
    identical. Measurement~2 is discordant because the three estimates
    disagree. The concordia age is the most likely age given the
    analytical uncertainties. It falls between the other two age
    estimates, and offers the best analytical precision of the
    three.}
  \label{fig:concordia}
\end{figure}

@@

The single grain concordia age combines the chronometric power of the
\textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb systems. For young
($<$1~Ga) samples, the concordia age is nearly identical to the
\textsuperscript{206}Pb/\textsuperscript{238}U age. For old samples
($>$2~Ga) it approaches the
\textsuperscript{207}Pb/\textsuperscript{206}Pb age. For Proterozoic
samples, the concordia age gradually shifts from the
\textsuperscript{206}Pb/\textsuperscript{238}U to the
\textsuperscript{207}Pb/\textsuperscript{206}Pb age. Using concordia
ages removes the need for an arbitrary cutoff between the two
chronometers. An additional advantage is that the concordia age offers
better precision than the
\textsuperscript{206}Pb/\textsuperscript{238}U and the
\textsuperscript{207}Pb/\textsuperscript{206}Pb chronometer (or the
\textsuperscript{207}Pb/\textsuperscript{235}U for that matter).
Figure~\ref{fig:precision} quantifies this effect using a
semi-analytical mass spectrometry simulation whose algorithm is
provided in Appendix~\ref{app:synthetics}.

\begin{figure}[t]
  \includegraphics[width=8.3cm]{agerr.pdf}
  \caption{Predicted uncertainties of the
    \textsuperscript{206}Pb/\textsuperscript{238}U ($t_{68}$),
    \textsuperscript{207}Pb/\textsuperscript{235}U ($t_{75}$),
    \textsuperscript{207}Pb/\textsuperscript{206}Pb ($t_{76}$) and
    concordia ($t_c$) ages for a synthetic dataset with a constant
    uranium concentration. Dwell times and detector sensitivities were
    chosen so as to yield results that are similar to those obtained
    from real data. The concordia age (solid line) always offers the
    best precision. See Appendix~\ref{app:synthetics} for further
    details about the calculations behind this figure.}
  \label{fig:precision}
\end{figure}

@@

The most common definition of discordance uses the relative difference
between the \textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb age estimate 
\citep{gehrels2011}:
\begin{equation}
  d_r = 1 - t_{68}/t_{76}
  \label{eq:dr}
\end{equation}

@@

Detrital zircon U--Pb data can be filtered by removing all
measurements whose discordance values exceed a certain threshold
value. Typical cutoff values for $d_r$ are 10--30\%
\citep{gehrels2011}, whereas $d_p$ is generally set to 5\%
\citep{spencer2016}. Different discordance criteria produce different
U--Pb age spectra. For example, a relative age cutoff will
preferentially remove young grains whereas an absolute age cutoff is
comparatively more likely to remove old grains
(Figure~\ref{fig:agediscordance}).

\begin{figure}[t]
  \includegraphics[width=7.3cm]{TW-option-1234.pdf}
  \caption{Discordance cutoffs for four of the six discordance
    definitions discussed in Sections~\ref{sec:discordance1} and
    \ref{sec:discordance2}. The $d_{p}$ and $d_c$ criteria are not
    shown because they depend on the analytical uncertainty of the
    measurements, which may vary between studies. The top panel shows
    cutoffs of $d_r=20\%$ (relative age filter, dashed line),
    $d_t=300$~Ma (absolute age filter, dash-dot line), $d_{sk}=2\%$
    (Stacey--Kramers filter, dotted line) and $d_{a}=15\%$ (Aitchison
    distance, solid line) on a Tera-Wasserburg concordia diagram,
    which is plotted in logarithmic space to provide a more balanced
    view of the old and young ends of the time scale. The bottom panel
    presents the \textsuperscript{207}Pb/\textsuperscript{206}Pb --
    \textsuperscript{206}Pb/\textsuperscript{238}U age discordance
    along this line. This shows that the $d_r$-criterion favours young
    samples and the $d_{sk}$-criterion old samples, whereas the
    $d_a$-criterion is most strict for samples of intermediate (i.e.,
    Proterozoic) age. }
  \label{fig:agediscordance}
\end{figure}

The p-value definition affects grains differently depending on their
analytical precision \citep{nemchin2005}. For example, consider a
1.5~Ga zircon that is $d_r=1\%$ discordant. If this grain were
analysed by LA-ICP-MS with an analytical precision of 2\%, say, then
it would pass the chi-square test and be accepted as being
concordant. However, if that same grain were analysed by TIMS with a
precision of 0.2\%, then the p-value criterion would reject it as
being discordant (Figure~\ref{fig:TIMSvsLAICPMS}). It seems
fundamentally unfair that an imprecise analytical method would be
favoured over a precise one. This is a pertinent problem because
technical innovations are increasing the precision of all analytical
approaches to U--Pb geochronology.  As precision improves, so does the
ability to detect ever small degrees of discordance. Using the p-value
criterion, there may come a time when no detrital zircon passes this
filter.

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{TIMSvsICPMS.pdf}
  \caption{Application of the flawed p-value discordance criterion
    to two synthetic measurements by TIMS (left) and LA-ICP-MS
    (right).  The precise TIMS measurement is labelled as discordant
    even though it plots closer to the concordia line than the
    imprecise LA-ICP-MS measurement, which is labelled as
    concordant.
  }
  \label{fig:TIMSvsLAICPMS}
\end{figure}

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{TW-sk.pdf}
    \caption{Using the \citet{stacey1975} common Pb model as a
      discordance criterion. This criterion assumes that the
      discordance is caused by linear mixing (hence, the linear scale
      of this Tera-Wasserburg plot) between radiogenic Pb
      (intersections of the mixing lines with concordia) and common Pb
      (intersection of the mixing lines with the vertical axis). The
      dashed line marks the 20\% ($= \delta_1/[\delta_1+\delta_2]$)
      discordance cutoff. This discordance filter, which must be
      applied \emph{before} making any actual common Pb correction, is
      more forgiving for young grains than it is for old grains. In
      this respect, it has the opposite effect of the relative age
      filter shown in Figure~\ref{fig:agediscordance}.  }
    \label{fig:SK}
\end{figure}

@@

There exists a vast body of statistical literature detailing the
theoretical and practical advantages of logratio analysis. A deeper
discussion of this topic falls outside the scope of this paper, but
the interested reader is referred to \citet{aitchison1986} and
\citet{pawlowsky2015} for further information. The Euclidean distance
between logratios is also known as the `Aitchison distance'.
Discordance can be redefined as the Aitchison distance from the
measured logratios to the concordia line. We here introduce two ways
to do so.  A first option is to simply measure the distance along a
perpendicular line to the concordia curve
(Figure~\ref{fig:aitchison}):
\begin{equation}
  d_{a} = dx(t_{68})
  \sin\!\left(\arctan\!\left[ \frac{dy(t_{76})}{dx(t_{68})} \right]
  \right)
  \label{eq:dperp}
\end{equation}

@@

This definition produces a parallel band around the concordia line in
logarithmic Tera-Wasserburg space. In contrast with $d_r$, $d_t$,
$d_{sk}$, the $d_a$ criterion is less strict at both the young and old
extremes of the geological timescale, and more strict during the
Proterozoic Era (Figure~\ref{fig:agediscordance}), when the U--Pb
method is most reliable.

@@

@@

The data come from field areas on all seven continents. They were
acquired by LA-ICP-MS, on either an Agilent 7700x (early samples) or
an Agilent 7900 (recent samples) instrument, which was coupled to a
New Wave NWR193 excimer laser. Analytical conditions varied slightly
between different samples but generally used a 25 or 35 micron laser
spot, Ple\v{s}ovice zircon as a primary standard \citep{slama2008}
and, in most cases, GJ-1 zircon as a secondary standard
\citep{jackson2004}. Data reduction was done with GLITTER
\citep{griffin2008}. The data were not subjected to any common Pb
correction or other filters, apart from a visual inspection to remove
the most extreme outliers ($<1\%$ of the data).

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{KDE.pdf}
  \caption{Four superimposed kernel density estimates (with 60~Ma
    bandwidth) for 10,000 unfiltered detrital zircon U--Pb
    dates. The \textsuperscript{207}Pb/\textsuperscript{235}U,
    \textsuperscript{206}Pb/\textsuperscript{238}U and concordia age
    spectra are similar.  However the KDE of the
    \textsuperscript{207}Pb/\textsuperscript{206}Pb data stands
    apart from the other three curves. It deviates both at the young
    end of the age spectrum (which it suppresses), and at the old
    end (which it inflates).
  }
  \label{fig:KDE}
\end{figure}

@@

\begin{figure*}[t]
  \includegraphics[width=12cm]{KDEs.pdf}
  \caption{Filtered U--Pb age spectra for the test data, removing the
    50\% most discordant grains according to the six discordance
    filters reviewed in this paper. Note how the concordia distance
    filter ($d_c$, fifth panel) exhibits the smallest difference
    between the four dating methods (including
    \textsuperscript{207}Pb/\textsuperscript{206}Pb), whereas the
    p-value filter ($d_p$, sixth panel) generates the greatest
    differences between them.  This observation suggests that the
    $d_c$ criterion is the best and the $d_p$ criterion the worst of
    the six methods. Also note how the relative age filter ($d_r$)
    strongly inflates the older half of the age spectrum at the
    expense of the younger half. The Stacey-Kramers filter ($d_{sk}$)
    has the opposite effect, as expected from
    Figure~\ref{fig:agediscordance}.}
  \label{fig:KDEs}
\end{figure*}

Figure~\ref{fig:KDEs} removed 50\% of the data, in order to emphasise
the differences between the six filters. In real applications, less
stringents discordance filters are usually applied. As mentioned in
the introduction, most current detrital zircon studies apply a
10\%--30\% relative age cutoff.  Using the test data, we can evaluate
the equivalent values for the $d_t$, $d_t$, $d_{sk}$, $d_a$ and $d_c$
criteria (Table~\ref{tab:conversiontable}). For example, a relative
age filter of 10\% removes the same fraction of the test data as an
absolute age filter with $d_t=83.5$~Ma, a Stacey-Kramers filter with
$d_{sk}=0.381$\%, a perpendicular Aitchison filter with $d_a=3.61$\%,
or a concordia distance filter with $d_c=3.67$\%.
